---
title: "generalized linear mixed models"
author: "Ben Bolker"
bibliography: "../glmm.bib"
---

```{r knitr,include=FALSE}
require("knitr")
knit_hooks$set(crop=hook_pdfcrop)
opts_chunk$set(fig.width=5,fig.height=4,
               out.width="0.6\\textwidth",
               fig.align="center",
               tidy=FALSE,echo=FALSE,warning=FALSE,message=FALSE)
```

```{r pkgs}
library(lattice)
library(plotrix) ## for axis.break
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(mvbutils) ## for foodweb()
library(lme4)
library(grid)
zmargin <- theme(panel.spacing=unit(0,"lines"))
library(scales) ## for 'scientific', 'trans_format'
theme_set(theme_bw())
```

\newcommand{\X}{\bm X}
\newcommand{\Z}{\bm Z}
\newcommand{\bbeta}{\bm \beta}
\newcommand{\bb}{\bm b}
\newcommand{\bu}{\bm u}
\newcommand{\bLambda}{\bm Lambda}
\newcommand{\bEta}{\bm \eta}
\newcommand{\btheta}{\bm \theta}
\newcommand{\bzero}{\bm 0}

## (Generalized) linear mixed models

(G)LMMs: a statistical modeling framework incorporating:

-   combinations of categorical and continuous predictors,\
    and interactions
-   (some) non-Normal responses\
    (e.g. binomial, Poisson, and extensions)
-   (some) nonlinearity\
    (e.g. logistic, exponential, hyperbolic)
-   non-independent (grouped) data

![image](pix/models-glmm.pdf){width="\textwidth"}

## Coral protection from seastars (*Culcita*) by symbionts [@mckeon_multiple_2012]

```{r culcita1}
stierdat <- read.csv("../data/culcitalogreg.csv")
stierdat$ttt <- as.factor(stierdat$ttt) ## treatment should be a factor
## contrast matrix
cmat <- matrix(c(0,1,1,1,0,1,-1,0,0,1,1,-2),ncol=3,
               dimnames=list(NULL,c("symb","crab.vs.shr","addsymb")))
contrasts(stierdat$ttt) <- cmat
stierdat$block <- as.factor(stierdat$block) ## not really necessary but logical
stierdat <- stierdat[,1:4]   ## don't really need the rest
mod0 <- glm(predation~ttt+block,binomial,data=stierdat)
mod1 <- glm(predation~ttt,binomial,data=stierdat)
library(lme4)
mod2 <- glmer(predation~ttt+(1|block),family=binomial,data=stierdat)
## AGQ 
mod2B <- glmer(predation~ttt+(1|block),family=binomial,
               nAGQ=20,data=stierdat)
library(MASS)
library(nlme)
mod5=glmmPQL(predation~ttt,random=~1|block,family=binomial,data=stierdat)
levels(stierdat$ttt) <- c("none","shrimp","crabs","both")
m <- melt(stierdat[,1:3],id.vars=1:2)
m2 <- dcast(m,ttt~variable,fun.aggregate=mean)
m3 <- dcast(m,ttt+block~variable,fun.aggregate=sum)
p <- with(m3,table(predation,ttt))
``` 

```{r bplot1,fig.width=8,fig.height=6,out.width="0.8\\textwidth"}
op <- par(las=1,cex=1.5,bty="l")
bb <- barplot(p,ylab="Number of blocks",xlab="Symbionts",
              main="Number of predation events")
bloc <- apply(p,2,cumsum)
midb <- rbind(bloc[1,]/2,
              (bloc[1,]+bloc[2,])/2,
              (bloc[2,]+bloc[3,])/2)
text(bb[col(p)][p>0],midb[p>0],c(0,1,2)[row(p)][p>0])
par(op)
```

## Environmental stress: *Glycera* cell survival (D. Julian unpubl.)

```{r glycera}
x <- read.csv("../data/Live-Dead Tabulated Counts.csv")
## utility function for factoring/renaming variables before
##  lattice plot
rnfac <- function(dat,vars) {
  if (!all(vars %in% names(dat))) stop("unknown variable")
  for (v in vars) {
    dat[[v]] <- factor(dat[[v]])
    levels(dat[[v]]) <- paste(v,"=",round(as.numeric(levels(dat[[v]])),2),sep="")
  }
  dat
}
sc <- function(x) { (x-min(x))/diff(range(x))}
xsc <- x
predvars <- c("Osm","Cu","H2S","Anoxia")
for (i in predvars) {
  xsc[[i]] <- sc(xsc[[i]])
}
xsc$Osm <- xsc$Osm-0.5
## xsc$O2 <- 1-xsc$O2
## names(xsc)[names(xsc)=="O2"] <- "anox"
xr0 <- within(x,FractionAlive <- Alive/(Alive+Dead))
xr <- melt(subset(xr0,select=-c(Alive,Dead)),id.vars=1:5)

x4 <- dcast(xr,H2S+Anoxia+Cu+Osm~.,fun.aggregate=mean)
names(x4)[5] <- "all"
x5 <- rnfac(x4,c("Anoxia","Osm"))

## FIXME: replace with ColorBrewer colours?
cmgen.colors  <- function (n,h1=6/12,h2=10/12,maxs=0.5)  {
    if ((n <- as.integer(n[1])) > 0) {
        even.n <- n%%2 == 0
        k <- n%/%2
        l1 <- k + 1 - even.n
        l2 <- n - k + even.n
        c(if (l1 > 0) hsv(h = h1,
                          s = seq(maxs, ifelse(even.n, 0.5/k, 0), length.out = l1),
                          v = 1),
          if (l2 > 1) hsv(h = h2,
                          s = seq(0, maxs, length.out = l2)[-1],
                          v = 1))
    }
    else character(0)
}
rb.colors <- function(n) {
  cmgen.colors(n,h1=0,h2=0.7,maxs=1)
}
```
 
## *Arabidopsis* response to fertilization & herbivory [@banta_comprehensive_2010]

## Coral demography (J.-S. White unpubl.)
```{r}
load("../data/culcita.RData")
library(lme4)
cmod1 <- glmer(predation~ttt+(1|block),data=culcita_dat,
               family=binomial,nAGQ=10)
cmod3 <- glm(predation~ttt,data=culcita_dat,family=binomial)
## jump through hoops to get GLM to estimate without complete-separation blocks 7,8,9 messing everything up!
cmod4 <- glm(predation~ttt+block,data=transform(culcita_dat,block=relevel(block,2)),
             family=binomial,
             start=c(2,-4,-4,-6,-2,0,0,4,5,30,30,30,5))
             ## contrasts=list(block=contr.sum),
             ## start=rep(0,)
             ## start=c(10,-4,-5,-6,-20,-8,-8,-8,-5,-3,20,20,20))
r <- ranef(cmod1,condVar=TRUE)
## set up plots
library(coefplot2)
f1 <- coeftab(cmod1)
f3 <- coeftab(cmod3)
f4 <- coeftab(cmod4)[1:4,]
r1 <- coeftab(cmod1,ptype="ranef")
r2 <- f3[1,]  ## intercept
r4 <- coeftab(cmod4)[-(1:4),]
v <- data.frame(predict(cmod4,
             newdata=data.frame(block=factor(1:10),ttt="none"),
             se.fit=TRUE))
v$fit <- v$fit-mean(v$fit)
v <- setNames(v[,1:2],c("Estimate","Std..Error"))
w <- data.frame(Estimate=rep(0,10),
                Std..Error=NA,row.names=1:10)
## less clunky way to do this?
dd <- function(X,method,type,
               strip="^ttt") {
  data.frame(method,type,
             p=gsub(strip,"",rownames(X)),X[,1:2],
             stringsAsFactors=FALSE)
}
allEsts <- rbind(dd(f1,"mixed","ttt"),
      dd(f3,"pooled","ttt"),
      dd(f4,"fixed","ttt"),
      dd(r1,"mixed","blocks",
         strip="block\\.\\(Intercept\\)"),
      dd(v,"fixed","blocks"),
      dd(w,"pooled","blocks"))
allEsts <- transform(allEsts,
   p=factor(p,levels=gtools::mixedsort(unique(p))))
```

```{r glycplot1,out.width="\\textwidth",fig.width=8,fig.height=5}
orig <- trellis.par.get()
pad <- 0 ## 15 for regular layout
trellis.par.set(layout.widths=list(right.padding=pad,left.padding=pad),
                regions=list(col=rb.colors(100)),
##                regions=list(col=brewer.pal(11,"RdBu")),
## leave text alone for regular layout
                add.text=list(cex=0.8),axis.text=list(cex=0.5))
levels(x5$Anoxia) <- c("Normoxia","Anoxia")
## print(levelplot(`(all)`~factor(H2S)*factor(Cu)|Anoxia*Osm,
##          col.region=rb.colors(100),
##          data=x5,
##          xlab=expression(H[2]*S),
##          ylab="Copper"))
levelplot(all~factor(H2S)*factor(Cu)|Osm*Anoxia,
                col.region=rb.colors(100), ## brewer.pal(11,"RdBu"), ## rb.colors(100),
                data=x5,
                xlab=expression(H[2]*S),
                ylab="Copper")
trellis.par.set(theme=orig) ## restore settings
## FIXME: redo in ggplot2?  LOW PRIORITY
```

```{r arabplot1,fig.width=7,fig.height=6,out.width="0.7\\textwidth"}
panel.stripplot2 <-
function (x, y, jitter.data = FALSE, factor = 0.5, amount = NULL, 
    horizontal = TRUE, groups = NULL, ...) 
{
    if (!any(is.finite(x) & is.finite(y))) 
        return()
    panel.sizeplot(x = x, y = y, jitter.x = jitter.data && !horizontal, 
        jitter.y = jitter.data && horizontal, factor = factor, 
        amount = amount, groups = groups, horizontal = horizontal, 
        ...)
}
load("../data/Banta.RData")
trellis.par.set(list(fontsize=list(text=20)))
stripplot(jltf ~ amd|nutrient, 
                data=within(dat.tf,jltf <-jitter(log(total.fruits+1),
                  amount=0.05)),
                strip=strip.custom(strip.names=c(TRUE,TRUE)),
                groups=gen, type=c('p','a'),
                ylab="Log(1+fruit set)")
##                main="panel: nutrient, color: genotype")
## trellis.par.set(theme=orig) ## restore settings
```
